---
title: "Modelling interactions and contrasts"
format:
  html:
    embed-resources: true
editor: visual
---

# Modelling interactions and contrasts

Spider leg hair pulling / pushing friction dataset (Wolff, Gorb (2013))

```{r}
library(tidyverse)
library(downloader)
library(rafalib)
```

```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider_wolff_gorb_2013.csv"
if (!file.exists(filename)) download(url, filename)
spider <- read.csv(filename, skip=1)
```

Data describes frictional coefficient for pulling / pushing, for each spider leg pair (L1, L2, L3, L4).

```{r}
head(spider)
```

# Experiment overview

Spiders have 4 leg pairs. Each leg pair is labelled L1 - L4. *Cupiennius. salei* spider tarsas and claw tuft were collected, air dried, then performed pulling / pushing simulations. Friction on each leg pair samples were measured via force transducer.

Friction coefficients were recorded for each leg pair during pulling, and pushing movement.

The overview of boxplot data depicts higher friction coefficients in pulling movements. Early interpretations: leg pair L4 has a notably wider range of friction coefficients for pushing movements, suggesting pushing movement (moving away/backwards/rotate) is predominantly performed with the back leg pair (L4).

```{r}
boxplot(spider$friction ~ spider$type*spider$leg, main = "Comparison of friction coefficients of different leg pairs", ylab = "Friction coefficient", xlab = "Spider leg type")
```

First column (intercept) = 1s (typepull), second column is half 0, half 1 (typepush, typepull respectively). I.e. dataset involves 2 friction coefficients: typepull, and typepush (baseline intercept value + typepush offset).

```{r}
spider.sub <- spider[spider$leg == "L1",] # filter L1 leg pair subset (rows, keep matrix) [r,] from spider dataset
X <- model.matrix(~type, data = spider.sub) # filter L1 pairs by pull / push types
colnames(X)
imagemat(X, main = "Model matrix for linear model with interactions")
```

```{r}
stripchart(split(spider.sub$friction, spider.sub$type), # plot friction coefficients for L1
           vertical = TRUE, pch = 1, method = "jitter", las = 2,
           xlim = c(0,3), ylim = c(0,2))
```

Above dataset was modelled with one variable for demonstration of interactions. (Linear models are typically performed with \>2 variables):

type = pull / push

leg = L1-L4

Generate intercept (typepull L1), typepush (for all leg pairs):

```{r}
# X <- model.matrix(~type, data = spider) # filters into pull (intercept) and push types for all leg types
X <- model.matrix(~type + leg, data = spider)  # filters
head(X)
imagemat(X, main = "Spider leg pair L1-L4 + type (pull/push) matrix linear modelling")
```

x: 1 = typepull (L1); 2 = typepush (L1); 3 = L2; 4 = L3; 5 = L4

grey = 1; white = 0

col3 (L2 leg pair) has a considerably lower range of values than col4 (L3), suggesting fewer range of friction coefficient data for L2 than the other leg pairs.

```{r}
sum(spider == "L2")
sum(spider == "L3")
sum(spider == "L4")
```

Fit the linear model: friction against leg pair + type (pull/push). Use + to simulate offset/additive effect.

```{r}
head(spider)
fit2 <- lm(friction ~ leg + type, data = spider)
summary(fit2)

spider$group <- factor(paste0(spider$leg, spider$type))
stripchart(split(spider$friction, spider$group), vertical = TRUE, method = "jitter", pch = 1, las = 2, xlim = c(0,9))
# pch = dot shape, las = x-axis value text orientation
```

Friction coefficient of each LegPair/Type is not equivalent to the mean - the above demonstration assumes equal variance across all leg pull/push groups.

Where we cannot assume equal variance, data must be transformed e.g. via log or sqrt, to reflect the variance, then use lm().

Or, use Welch's t-test or Wilcoxon rank sum for unequal variance data, without needing to transform values.

```{r}
coef <- coef(fit2)
coef
```

The above coefficients compare each group to the intercept (L1 pull).

# Interactions

The above compares coefficient for each leg pair against the baseline reference (L1)). The above demonstration uses coefficient leg + type. The + acts as a fixed offset, to add a universal typepush coefficient, to add onto each leg pull coefficient.

Leg \* type describes the interaction between each group: the effect of typepush differs for each leg pair. This introduces new typepush coefficients for L2, L3, L4, instead of using L1 typepush as a universal/fixed offset for leg pushing.

```{r}
X <- model.matrix(~ type*leg, data = spider)
colnames(X)
imagemat(X, main = "Model matrix for linear models with interactions")
```

Fit the new X model matrix to linear model lm() to reflect typepush of each leg pair:

```{r}
fit3 <- lm(friction ~ type*leg, data = spider)
summary(fit3)
coefs <- coef(fit3)
```

P values compare significance between L2 typepush vs L1 typepush (reference).Ltypepush:legL2 has p \> 0.05. However, typepush:legL3 vs L1 typepush has significant p \< 0.05.

To compare coefficients between groups L2 and L3, use contrast(): subtract L3 coefficient from L2:

```{r}
library(contrast)
L3vsL2 <- contrast(fit2, list(leg = "L3", type = "pull"), list(leg = "L2", type = "pull"))
L3vsL2

# or, equivalent to (produces same value as contrast():
coefL3vsL2 <- coef[3] - coef[2]
coefL3vsL2
```

Contrast = change from L3 to L2 = -0.1142949 / decrease in friction coefficient

To compare push vs pull coefficients for same leg pair (L2):

```{r}
L2push.vs.pull <- contrast(fit3,
                           list(leg = "L2", type = "push"), list(leg = "L2", type = "pull"))
L2push.vs.pull
C <- L2push.vs.pull$X # open matrix
C
# above contrast() is also equivalent to:
coefs[2] + coefs[6]
```

==\> typepush = -0.618,

i.e. typepush for L2 is 0.618 less in friction coefficient than L2 typepull

## Contrasts: difference of differences

Above compares between 2 variables: type (push or pull) for a leg pair (L1-L4), vs another type(push or pull) for a leg pair (L1-L4). e.g. L2 typepush vs L1 typepush.

It is difficult to compare differences between leg pairs using the p-values from fit3, as these are based on comparison between that respective leg pair **against baseline (L1)**.

To compare difference of differences, use ANOVA: analysis of variance: analyse general variance in push ***and*** pull against other leg pairs. What is the likelihood that the differences in typepull and typepush across all leg pairs (L1-L4) are due to chance?

```{r}
anova(fit3) # produces F-test: testing 3 coefficients at once (df)
```

type \< 0.05, significant p-value (also based on F-value): i.e. there is a significant difference between typepush and typepull across all leg pairs.

i.e. spiders have significantly higher friction for pulling movement vs pushing movement.

## Interactions: Practice questions

Suppose we have an experiment with two species A and B, and two conditions, control and treated.

```{r}
species <- factor(c("A","A","B","B"))
condition <- factor(c("control","treated","control","treated"))
```

We will use the formula of `~ species + condition` to create the model matrix:

```{r}
model.matrix(~ species + condition)
# 1 = speciesA control [1,0,0]; 2 = speciesA treated [1,0,1]; 3 = (speciesA) + speciesB control [1,1,0]; 4 = (speciesA) + speciesB treated [1,1,1]
```

What should the contrast vector be, to obtain the difference between the species B control group and the species A treatment group (species B control - species A treatment)? Assume that the coefficients (columns of design matrix) are: Intercept, speciesB, conditiontreated.

\[1,1,0\] - \[1,0,1\] = \[0,1,-1\] (Ans = D)

Use the Rmd script to load the spider dataset. Suppose we build a model using two variables: \~ type+ leg.

What is the t-statistic for the contrast of leg pair L4 vs. leg pair L2?

```{r}
# model.matrix(~type + leg, data = spider)
S <- lm(friction ~ type + leg, data = spider)
S
library(contrast)
contrast(S, list(leg = "L4", type = "pull"), list(leg = "L2", type = "pull"))
# contrast = 0.1094167
# t-statistics = 2.45
```

The t-statistic for the contrast of leg pair L4 vs. leg pair L2 is constructed by taking the difference of the estimated coefficients legL4 and legL2, and then dividing by the standard error of the difference.

Recap:

Calculate betahat via observed values (Y) and matrix (X) (data values).

Predict Y observed outcomes (Yhat) by fitting betahat: Y = XBeta (+ error) –\> generate a regression line.

Covariance measures the shared error between two variables (of the same dataset here), i.e. if they are subject to the same error, it would not independently affect the difference,

Hence why var(betahatL4 - betahatL2) = var(betahatL4) + var(betahatL2) - 2cov(betahatL4, betahatL2)

```{r}
X <- model.matrix(~ type + leg, data = spider)
```

Suppose that we notice that the within-group variances for the groups with *smaller frictional coefficients are generally smaller*, and so we try to apply a transformation to the frictional coefficients to make the within-group variances more constant.

```{r}
spider$log2friction <- log2(spider$friction) # add log2friction column to spider dataset
boxplot(log2friction ~ type*leg, data = spider)

# Run a linear model of log2friction with type, leg and interactions between type and leg:
log2 <- lm(log2friction ~ type*leg, data = spider)

# What is the t-statistic for the interaction of type push and leg L4? If this t-statistic is sufficiently large, we would reject the null hypothesis that the push vs. pull effect on log2(friction) is the same in L4 as in L1.
summary(log2)
# t-statistic = -3.689 (SE away from mean)

# very large t value -12.901 for typepush (L1), i.e. suggesting a significant difference in push vs pull (L1)

# likewise, L4 also has significant push vs pull difference (-3.689). However, not same as L1 difference push vs pull, hence reject null.

```

Using the same analysis of log2 transformed data, What is the F-value for all of the `type:leg` interaction terms in an ANOVA? If this value is sufficiently large, we would reject the null hypothesis that the push vs. pull effect on `log2(friction)` is the same for all leg pairs.

```{r}
aov <- anova(log2) # includes interaction variables type*leg
summary(aov)
aov
# anova tests: does adding these interaction variables *actually reduce variance? (How likely is it that this reduction in variance is due to chance?)

# F-statistic = 10.701 for type:leg (i.e. 10x stronger variance than the noise, unlikely variance is due to chance)

```

What is the L2 vs L1 estimate in log2(friction) for the pull samples?

```{r}
contrast(log2, list(leg = "L2", type = "pull"), list(leg = "L1", type = "pull"))
# t value = 2.9 (>2), p value = 0.004; significant differnece in L1 vs L2 pull friction coefficients
# *this could also be seen simply from summary(log2): legL2 = 0.34861
```

What is the L2 vs. L1 estimate in `log2(friction)` for the push samples? Remember, because of the interaction terms, this is not the same as the L2 vs L1 difference for the pull samples. If you’re not sure use the `contrast` function. Another hint: consider the arrows plot for the model with interactions.

```{r}
contrast(log2, list(leg = "L2", type = "push"), list(leg = "L1", type = "push"))
```
